Phylogram instead of chronogram when assessing the neutral evolution of a trait

### 2025-11-17 manuscript/main.tex
Formalism rational draft

\newcommand{\Multiply}{\cdot}
\DeclareMathOperator{\Var}{\text{var}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\PhyloSupport}{\pi}
\newcommand{\MeanPhyloSupport}{\overline{\PhyloSupport}}
\newcommand{\Indiv}{k}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\MutationRatePheno}{\mu}
\newcommand{\MutationRateNuc}{u}
\newcommand{\NbrLoci}{L}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\brownian}{\mathcal{B}}

@@

\subsection*{Brownian motion (BM)}
On the simulated dataset, we fitted a single rate BM using either a phylogram or a chronogram as the underlying tree.
We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian motion (BM) on continuous traits evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
The data consist of the mean trait value for each extant species, and the tree topology is fixed.
Along each branch, the value of the trait is drawn from a normal distribution with mean equal to the parent node value and variance equal to the branch length~\citep{felsenstein_phylogenies_1985}.
Formally, the BM process is described by the equation:
% Ornstein-Uhlenbeck model equations
\begin{align}
    \der \Trait_t & = \sigma \Multiply \der W_t,  \label{eq:bm-process}
\end{align}
where $\Trait_t$ is the trait value at time $t$, $\sigma$ is the rate of the BM, and $\der W_t$ is a Wiener process (i.e. a standard Brownian motion).
We used a log-uniform prior for $\sigma$.
Using Markov chain Monte Carlo (MCMC), we reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.
For traits simulated under neutral evolution, we expect more accurate trait reconstruction on a phylogram, and conversely, for traits under selection (moving optimum), more accurate trait reconstruction on a chronogram.

\subsection*{BM with a switch}
To test the support of a phylogram over a chronogram for the simulated and empirical datasets, we implemented a model based on a single rate BM.
The model was implemented in \textit{RevBayes} and contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
The tree topology is fixed, meaning both trees have the same branching structure, but because branch lengths are different and not necessarily on the same scale, both trees are re-scaled by dividing branch lengths by the total tree length.
As in the previous section, a BM is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability that the phylogram is favoured over the chronogram.
We expect $\PhyloSupport = 1$ for a trait under neutral evolution, and $\PhyloSupport = 0$ for a trait under selection (moving optimum).

\subsection*{Alternative models of evolution}
In addition to a single rate BM, we  also fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
For each branch, we draw a rate-multiplier either equal to $1$ (no rate shift), or drawn from a log-normal distribution with a median of 1, and a standard deviation such that rate shifts range over about one order of magnitude.
We set to our prior expected number of rate shifts to $1$ across the whole tree.
The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1000 gen., running of 50000 gen., 2 chains).

% Simple OU model versus a brownian process.
Also, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
% Ornstein-Uhlenbeck model equations
The OU process is described by the equation:
\begin{align}
    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \Multiply \der W_t, \label{eq:ou-process}
\end{align}
where the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, $\theta$ is the optimum value and the other parameters are defined as in eq~\ref{eq:bm-process}.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the BM, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
% alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
The prior for $\theta$ is uniform between the minimum and maximum trait value for extant species.
The prior for $\alpha$ is a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution with mean equal to half the root age divided by $\ln (2)$, meaning that we expect a phylogenetic half life of half the tree age.
Thus, when $\alpha = 0$ the OU process (eq.~\ref{eq:ou-process}) is equivalent to a BM (eq.~\ref{eq:bm-process}).
The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).

% Relaxed OU model, how many optimums?
Finally, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).
For each branch, we draw a shift in optimum either equal to $0$ (no shift), or drawn from a uniform distribution centred on $0$ with breadth spanning the whole range of observed trait value.
We set to our prior expected number of optimum shifts to $1$ across the whole tree.

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_pipeline}\\
    \vspace{10pt}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small \textbf{C.\ Simulations with constant generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_cstGT_phylogram_support}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small \textbf{D.\ Simulations with fluctuating generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_phylogram_support}
    \end{minipage}
    \caption{
        \textbf{Testing the support for a phylogram over a chronogram}
        Panel~A:
        Test for the support of a phylogram over a chronogram using a Brownian motion (BM) to model trait changes along a phylogeny.
        The model contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time ($\PhyloSupport=0$, chronogram in purple) to units of nucleotide divergence for neutral sites ($\PhyloSupport=1$, phylogram in green).
        The posterior mean estimate of $\PhyloSupport$ indicates whether a BM supports better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given the mean trait value for each extant species and both trees (chronogram \& phylogram) with the same topology.
        Only if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$) there is some support that the trait might evolve under drift.
        Panel~B: Wright-Fisher simulations of trait evolution along a phylogeny.
        Neutral evolution is modelled as a constant fitness regardless of the phenotype of individuals (yellow).
        Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree: as a moving optimum (blue) or as multiple optima (red).
        At each node of the tree we have a speciation event and the two descendant species evolve independently.
        The output of the simulation is the mean trait value for each extant species, but we also save the phylogram and the chronogram as obtained from independent markers.
        We then compute the posterior mean estimate of $\PhyloSupport$ from the output of the simulation.
        Panels C \& D: Violin plot of the posterior mean of $\PhyloSupport$ across 100 replicates for the different simulated regimes of evolution with mean $\PhyloSupport$ values at the center ($\MeanPhyloSupport$).
        Horizontal lines inside the violins are the estimated $\PhyloSupport$ of each replicate simulations.
        Results of simulations with constant generation time (Panel~C) and with fluctuating generation time, mutation rate and effective population size (Panel~D).
    }
    \label{fig:results-brownian}
\end{figure*}
%TC:endignore

@@

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We first tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
We simulated traits evolving on a phylogenetic tree under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum, and 3) under multiple optima (see Methods, Fig.~\ref{fig:results-brownian}A-C).
We first assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram, under the assumption that changes follow a Brownian motion (BM).
For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).

We then tested for the BM support of phylogram over a chronogram, and we developed as statistic denoted as $\PhyloSupport$, which ranges from 0 (full support for the chronogram) to 1 (full support for the phylogram).
For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch lengths are equivalent in both trees.
Next, we simulated changing generation time, mutation rate (per generation) and effective population size ($\Ne$) along the phylogenetic tree, mimicking a mammalian range of changes.
In this case, for a simulated neutral trait, the BM was better fitted on a phylogram than on a chronogram, with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold (Fig.~\ref{fig:results-brownian}E, yellow).
Conversely, for a simulated trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}E, blue) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}E, red).

@@

% Mainly testing BM versus OU
Moreover, we tested alternatives to the BM with Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$), an expected result since the BM with a switch supported the chronogram.
For a trait evolving neutrally, $p_{\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.0\times 10^{-8}$), an unexpected result since the BM with a switch supported the phylogram, thus showing a high rate of mis-classification on a phylogram.
When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram, but not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram but not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
In summary, fitting an OU process on a phylogram results in biases not only for a trait under selection but also for a trait under neutral evolution.

@@

    \caption{
        \textbf{Mis-classification of trait evolution on a phylogram.}
        Violin plot of posterior parameter estimates for different models of trait evolution and different simulated regimes of selection.
        Horizontal lines inside the violin show the results of each replicate simulation.
        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow) and under a moving optimum (blue).
        Wilcoxon rank-test are performed between the paired estimates on the phylogram and the chronogram.
        Panel~A \& B: Fit of a relaxed Brownian motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
        Posterior estimate for the variance in rate parameters (Panel~A) and number of rate changes (Panel~B).
        Panel~C: Relative fit an Ornstein-Uhlenbeck (OU) process compared to BM, using either a phylogram or a chronogram.
        Posteriors estimates for the support of the OU process over the BM (reversible-jump).
        Panel~D: Fit of a relaxed OU with multiple optima, using either a phylogram or a chronogram.
        Posterior estimate for the number of optimum changes.
    }
    \label{fig:results-alternative}
\end{figure*}
%TC:endignore

\section*{Discussion}
% Context and rationale
In phylogenetic comparative methods, the evolution of continuous traits is typically modelled as a stochastic process running along the branches of a phylogenetic tree.
The null model of neutral trait evolution, genetic drift, is thought of as a Brownian motion (BM) running on the tree~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
As a result, a trait on which the BM is favoured over alternative models are sometimes interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
However, the BM is not only the null model of neutral evolution, but can also model selection, with a trait tracking a moving optimum, where the optimum itself is following a BM~\citep{hansen_translating_1996, latrille_detecting_2024}.
Distinguishing these scenarios is not trivial, but we here showed that the underlying backbone tree to model trait evolution can help disentangle neutral evolution from selection.
The standard in phylogenetic comparative methods is a chronogram, where the branch lengths are measured in time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}, while phylograms, where the branch lengths are measured in number of substitutions instead, have been generally overlooked.
For a simulated trait under selection we assessed that the BM has indeed a better fit on a chronogram than on a phylogram.
However, we argue that to model a trait evolving neutrally, the backbone tree should not be a chronogram but instead a phylogram.
Phylograms would better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
In practice, we show that for simulations of a trait evolving neutrally, the fit of a BM supports a phylogram rather than a chronogram.
Additionally, for a neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was more accurate on a phylogram.
Altogether, the combined use of a phylogram and a chronogram can help support the hypothesis of neutral evolution, or to rule it out.
% Thus, we argue that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the BM under both a phylogram and a chronogram as the underlying tree.

% When does it apply and why is it important
We applied this method to different empirical datasets of body and brain masses in mammals, and we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these traits~\citep{latrille_detecting_2024}.
The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree (e.g. from 6 months for mice to 50 years for bowhead whales).
But more generally, our method is also relevant when changes in mutation rate (per generation) and effective population size ($\Ne$) are expected.
First, normalizing by nucleotide divergence also accounts theoretically for variations in effective population size ($\Ne$) since the substitution rate of neutral mutations is equal to the mutation rate, such that $\Ne$ has no effect on the rate of neutral evolution~\citep{kimura_evolutionary_1968, ohta_population_1972}.
Moreover, if population structure were to impact the probability of fixation for neutral mutations (which would also impact a neutral trait), the phylogram would automatically absorb these changes.
This argument was partially tested in our simulation by modelling changes in both $\Ne$ and generation time (although not modelling population structure \textit{per se}), and the phylogram was favoured over the chronogram for a trait evolving neutrally.
Second, under certain conditions, the use of phylograms can also absorbs changes in mutation rate (per loci per generation), which in mammals can vary up to 10 fold~\citep{bergeron_evolution_2023}.
As in the case of our simulations, we assumed that the genetic architecture of the trait is constant and the mutation rate is homogeneous across the genome.
However, if the mutation rate would only change for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.
Finally, because chronogram are usually derived from phylogram by calibrating the tree with molecular clocks, using phylograms directly has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.

% When does it fall appart, our second result
These examples show that while phylograms are useful to model neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
In our simulations, we found that generally using a phylogram tends to favour the OU process over the BM model.
This suggests that trees in units of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
We recommend continuing to use chronograms for modelling selection acting on a trait, as they generally provide more accurate and reliable results.
Moreover, for a trait under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
Therefore, we argue that while phylograms can offer valuable insights~\citep{wilson_chronogram_2022}, they should be used with caution and in conjunction with chronograms to provide a more comprehensive understanding of trait evolution.
Contrarily to our cautious note, \citet{wilson_chronogram_2022} suggested that phylograms are more robust for studying discrete character evolution than chronogram.
This apparent contradiction is due to the different units of measurement used in the studies, as they measure branch lengths in units of morphological distance rather than substitutions per site~\citep{wilson_chronogram_2022}.
We thus argue that the use of a phylogram, measured in units of substitutions, could also be useful for studying the evolution of discrete characters, but the same caution should be applied.

>>>>

\newcommand{\Multiply}{\cdot}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{var}}
\DeclareMathOperator{\Cov}{\text{cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\PhyloSupport}{\pi}
\newcommand{\MeanPhyloSupport}{\overline{\PhyloSupport}}
\newcommand{\Indiv}{k}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\MutationRatePheno}{\mu}
\newcommand{\MutationRateNuc}{u}
\newcommand{\NbrLoci}{L}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\brownian}{\mathcal{B}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{T}
\newcommand{\NbrGen}{g_{\Spi, \Spj}}
\newcommand{\NucDiv}{d_{\Spi, \Spj}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\RootTrait}{\phi}
\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\SubRate}{q}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\sigma}^2_{\mathrm{B}}}
\newcommand{\EstRateWhithin}{\widehat{\sigma}^2_{\mathrm{W}}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_pipeline}\\
    \vspace{10pt}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small \textbf{C.\ Simulations with constant generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_cstGT_phylogram_support}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small \textbf{D.\ Simulations with fluctuating generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_phylogram_support}
    \end{minipage}
    \caption{
        \textbf{Testing the support for a phylogram over a chronogram}
        Panel~A:
        Brownian motion (BM) is used to model trait changes along a phylogeny, given a mean trait value for extant species.
        The model contains a switch variable, denoted $\PhyloSupport$, that switches the branch lengths in units of time ($\PhyloSupport=0$, chronogram) to units of nucleotide divergence for neutral sites ($\PhyloSupport=1$, phylogram).
        The posterior mean estimate of $\PhyloSupport$ indicates whether a BM supports better the phylogram over the chronogram ($0 \leq \PhyloSupport \leq 1$).
        Panel~B: Wright-Fisher simulations of trait evolution along a phylogeny.
        At each node of the tree a speciation event creates two descendant species evolving independently.
        Neutral evolution is modelled as a constant fitness regardless of the phenotype of individuals (yellow).
        Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree: as a moving optimum (blue) or as multiple optima (red).
        The output of the simulation is: 1) the mean trait value for each extant species, 2) the phylogram as obtained from independent neutral markers and 3) the chronogram derived from the phylogram assuming a relaxed molecular clock.
        The posterior mean estimate of $\PhyloSupport$ is inferred from the output of the simulation.
        Panels C \& D: Violin plot of the posterior mean of $\PhyloSupport$ across 100 replicates for the different simulated regimes of evolution with mean $\PhyloSupport$ values at the center ($\MeanPhyloSupport$).
        Horizontal lines inside the violins are the estimated $\PhyloSupport$ of each replicate simulations.
        Results of simulations with constant generation time (Panel~C) and with fluctuating generation time, mutation rate and effective population size (Panel~D).
    }
    \label{fig:results-brownian}
\end{figure*}
%TC:endignore

@@

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
We simulated traits evolving on a phylogenetic tree under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum, and 3) under multiple optima (see Methods).
We first assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram, under the assumption that changes follow a BM.
For a neutral trait, ancestral trait reconstruction was more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).

We then tested for the BM support of phylogram over a chronogram, and we developed a statistic denoted as $\PhyloSupport$ (Fig.~\ref{fig:results-brownian}A), which ranges from 0 (full support for the chronogram) to 1 (full support for the phylogram).
For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}B-C), which is expected since the branch lengths are equivalent in both trees.
Next, we simulated changing generation time, mutation rate (per generation) and effective population size ($\Ne$) along the phylogenetic tree, mimicking a mammalian range of changes.
In this case, for a simulated neutral trait, the BM was better fitted on a phylogram than on a chronogram, with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80$ of the $100$ replicates above the $0.95$ threshold (Fig.~\ref{fig:results-brownian}D, yellow).
Conversely, for a simulated trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69$ out of $100$ replicates below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}D, blue) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67$ of the $100$ below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}D, red).

@@

% Mainly testing BM versus OU
Moreover, we tested alternatives to the BM with OU models: with a single optimum and with multiple optima (see Methods).
First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$), an expected result since the BM with a switch supported the chronogram.
For a trait evolving neutrally, $p_{\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.0\times 10^{-8}$), an unexpected result since the BM with a switch supported the phylogram, thus showing a high rate of mis-classification on a phylogram.
When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram, but not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram but not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
In summary, fitting an OU process on a phylogram results in biases not only for a trait under selection but also for a trait under neutral evolution.

@@

    \caption{
        \textbf{Mis-classification of trait evolution on a phylogram.}
        Violin plot of posterior parameter estimates for different models of trait evolution and different simulated regimes of selection.
        Horizontal lines inside the violin show the results of each replicate simulation.
        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow) and under a moving optimum (blue).
        Wilcoxon rank-test are performed between the paired estimates on the phylogram and the chronogram.
        Panel~A \& B: Fit of a relaxed Brownian motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
        Posterior estimate for the variance in rate parameters (Panel~A) and number of rate changes (Panel~B).
        Panel~C: Relative fit of an Ornstein-Uhlenbeck (OU) process compared to BM, using either a phylogram or a chronogram.
        Posterior estimates for the support of the OU process over the BM (reversible-jump).
        Panel~D: Fit of a relaxed OU with multiple optima, using either a phylogram or a chronogram.
        Posterior estimate for the number of optimum changes.
    }
    \label{fig:results-alternative}
\end{figure*}
%TC:endignore

\section*{Discussion}
% Context and rationale
In phylogenetic comparative methods, the evolution of continuous traits is typically modelled as a stochastic process running along the branches of a phylogenetic tree.
The null model of neutral trait evolution, genetic drift, is thought of as a Brownian motion (BM) running on the tree~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
As a result, a trait on which the BM is favoured over alternative models are sometimes interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
However, the BM is not only the null model of neutral evolution, but can also model selection, with a trait tracking a moving optimum, where the optimum itself is following a BM~\citep{hansen_translating_1996, latrille_detecting_2024}.
Distinguishing these scenarios is not trivial, but we here showed that the underlying backbone tree to model trait evolution can help disentangle neutral evolution from selection.
The standard in phylogenetic comparative methods is a chronogram, where the branch lengths are measured in time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}, while phylograms, where the branch lengths are measured in number of substitutions instead, have been generally overlooked.
For a simulated trait under selection we assessed that the BM has indeed a better fit on a chronogram than on a phylogram.
However, we argue that to model a trait evolving neutrally, the backbone tree should not be a chronogram but instead a phylogram.
Phylograms better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
In practice, we show that for simulations of a trait evolving neutrally, the fit of a BM supports a phylogram rather than a chronogram.
Additionally, for a neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was more accurate on a phylogram.
Altogether, the combined use of a phylogram and a chronogram can help support the hypothesis of neutral evolution, or to rule it out.
% Thus, we argue that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the BM under both a phylogram and a chronogram as the underlying tree.

% When does it apply and why is it important
We applied this method to different empirical datasets of body and brain masses in mammals, and we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these traits~\citep{latrille_detecting_2024}.
The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree, from 6 months for mice to 50 years for bowhead whales~\citep{demagalhaes_database_2009, nielsen_eye_2016}.
But more generally, our method is also relevant when changes in mutation rate (per generation) and effective population size ($\Ne$) are expected.
First, normalizing by nucleotide divergence also accounts theoretically for variations in effective population size ($\Ne$) since the substitution rate of neutral mutations is equal to the mutation rate, such that $\Ne$ has no effect on the rate of neutral evolution~\citep{kimura_evolutionary_1968, ohta_population_1972}.
Moreover, if population structure were to impact the probability of fixation for neutral mutations (which would also impact a neutral trait), the phylogram would automatically absorb these changes.
This argument was partially tested in our simulation by modelling changes in both $\Ne$ and generation time (although not modelling population structure \textit{per se}), and the phylogram was favoured over the chronogram for a trait evolving neutrally.
Second, under certain conditions, the use of phylograms can also absorb changes in mutation rate (per loci per generation), which in mammals can vary up to 10 fold~\citep{bergeron_evolution_2023}.
As in the case of our simulations, we assumed that the genetic architecture of the trait is constant and the mutation rate is homogeneous across the genome.
However, if the mutation rate would only change for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.
Finally, because chronograms are usually derived from phylograms by calibrating the tree with molecular clocks, using phylograms directly has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.

% When does it fall appart, our second result
These examples show that while phylograms are useful to model neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
In our simulations, we generally found that using a phylogram tends to favour the OU process over BM.
This suggests that trees in units of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
We recommend continuing to use chronograms for modelling selection acting on a trait, as they generally provide more accurate and reliable results.
Moreover, for a trait under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
Therefore, we argue that while phylograms can offer valuable insights~\citep{wilson_chronogram_2022}, they should be used with caution and in conjunction with chronograms to provide a more comprehensive understanding of trait evolution.
Contrarily to our cautious note, \citet{wilson_chronogram_2022} suggested that phylograms are more robust for studying discrete character evolution than chronogram.
This apparent contradiction is due to the different units of measurement used in the studies, as they measure branch lengths in units of morphological distance rather than substitutions per site~\citep{wilson_chronogram_2022}.
We thus argue that the use of a phylogram, measured in units of substitutions, could also be useful for studying the evolution of discrete characters, but the same caution should be applied.
